Aquifer depletion exacerbates agricultural drought losses in the US High Plains

### 2023-11-01 Writing/manuscript.Rnw
updates on manuscript writing (mostly on data)

Meteorological data needed to estimate spatial and temporal variations in drought conditions across the HPA were obtained from the gridMET dataset \citep{Abatzoglou2013}. gridMET is a gridded daily weather dataset with a spatial resolution of \textasciitilde4km, and has been used in a range of climate impacts studies for agriculture and other sectors in the United States \citep{abatzoglou2016impact, pereira2015crop, crane2018machine, venkatappa2021impacts, zhu2019dissecting}. Daily precipitation values were extracted for each gridMET cell overlapping counties in our sample. Daily values of maximum and minimum temperature, surface radiation, wind speed, and humidity were also obtained from gridMET in order to calculate daily reference crop evapotranspiration for each grid cell. From these data, we computed the seasonal water deficit (\(D_{i,t}\)) for each grid cell by subtracting total precipitation from total reference evapotranspiration for the period May through September, which represents the typical growing season for corn and soybean in the HPA region. Seasonal water balance estimates were subsequently aggregated to the county-level to match our agricultural production data using an area-weighted average based on the proportion of the county covered by each overlapping gridMET cell. Figure \ref{fig:deficit-yield-hist} presents annual water deficit by year for the corn and soybean data, illustrating major drought events in the region in years such as 2000, 2002, and 2012. The importance of irrigation is clearly evident in years with large water deficits, with rainfed yields dipping substantially whereas irrigated yields are, on average, largely unaffected (Figure \ref{fig:deficit-yield-hist}).

@@

where county and year are indicated by subscripts \(i\) and \(t\), respectively, \(Y\) is crop yield, and \(D\) is seasonal water deficit. \(C_j\) is a dummy variable that takes 1 if observation \(i\) is under category \(C_j\), 0 otherwise. The observations are grouped into four categories: rainfed production, and county-years with aquifer thickness falling in the \([0\%-33.3\%)\), \([33.3\%-66.7\%)\), and \([66.7\% - 100\%]\) quantile ranges of aquifer thickness. The three irrigated production groups have about the same number of observations, which ensures statistical stability in estimating yield response to drought by category. Spline functions are used to capture the potential non-linearity of the impact of water deficit on yield in a flexible manner, with \(\sigma_j\) representing the category-specific intercept, \(\phi_{j,k}\) \(k\)th spline basis function, and \(\beta_k\) is its coefficient (see Appendix \ref{spline-basis} for further explanations on how spline-based regression works). $K$ is the total number of spline basis functions. $K = 3$ was chosen for this regression as it minimizes Bayesian Information Criterion (BIC). To control for time-invariant heterogeneity across counties and yearly shocks, county fixed effects (\(\alpha_i\)) and year fixed effects (\(\lambda_t\)) are also included and standard error estimation is clustered by county.

@@

where \(s_{i,t}\) is the share of irrigated production. The impact of average water balance is captured by \(\sum_{k}^K \eta_{k}\phi_{k}(D_{i})\), where \(\phi_{k}(\cdot)\) is the \(k\)th spline basis function and \(\eta_{k}\) is its coefficient. The impact of aquifer thickness is captured in a similar way with \(\tau_{l}(\cdot)\) and \(\beta_{l}\) being the \(l\)th spline basis function and its coefficient, respectively. $K$ and $L$ are the numbers of basis functions for $D$ and $A$, respectively. $K = 4$ and $L = 4$ were chosen as they minimize BIC. County-level soil characteristics including sand percentage, silt percentage, and water holding capacity, are represented by \(X_i\). State-year fixed effects and error term are represented by \(\theta_{s,t}\) and \(v_{i,t}\), respectively. Unlike for our regression models of per-area yields, we do not include county fixed effects to avoid eliminating the majority of variation in the aquifer thickness variable. The vast majority of variation in aquifer thickness is cross-sectional (across counties), rather than over time. Including county fixed effects would effectively eliminate the cross-sectional variation in aquifer thickness. A formal discussion of the trade-offs associated with the choice of fixed effects is provided in Appendix \ref{county-fe}.

@@

All the raw data-sets used in this study are publicly accessible. The boundary shape file of the High Plains aquifer was obtained from the U.S. Geological Survey at \url{https://water.usgs.gov/GIS/metadata/usgswrd/XML/ds543.xml#stdorder}. This data is used for Figure \ref{fig:sat-map}. Corn and soybean yield data were obtained from the U.S. Department of Agriculture's National Agricultural Statistics Service (USDA-NASS), available at \url{https://www.nass.usda.gov/Quick_Stats/}. The R tidyUSDA package \citep{RtidyUSDA} was used to automatically download the relevant data. This data is used for Figure \ref{fig:deficit-yield-hist} and the regression analysis. Weather data (precipitation, temperature, and evapotranspiration) was obtained from the gridMET dataset \citep{Abatzoglou2013}, which is available at \url{https://www.climatologylab.org/gridmet.html}. This data is used for Figure \ref{fig:deficit-yield-hist} and the regression analysis. The aquifer thickness data were obtained from Hydroshare \citep{haacker2023} available at \url{https://www.hydroshare.org/resource/7d925c7944244032af98c9ed20c22db6/}. This data is used for Figure \ref{fig:sat-map} and the regression analysis. Soil characteristics data was obtained from the Soil Survey Geographic Dataset, which is used for regression analysis. The R soilDB package \citep{Rsoildb} was used to download the data. Regression analyses are implemented using the R fixest package \citep{Rfixest}. Other key R packages include ggplot2 \citep{Rggplot2} for creating plots, data.table \citep{Rdatatable} for data wrangling, and sf \citep{Rsf} for spatial data operations. The data used for the regression analysis can be generated using our codes by following the instructions available at the linked Github repository.

\subsection{Code Availability}
All the codes used for this study are available at the following Github repository (\href{https://github.com/tmieno2/Drought-Production-Risk-Aquifer}{link}). All analyses were performed in R \citep{R}.

>>>>

Meteorological data needed to estimate spatial and temporal variations in drought conditions across the HPA were obtained from the gridMET dataset \citep{Abatzoglou2013}. gridMET is a gridded daily weather dataset with a spatial resolution of \textasciitilde 4km, and has been used in a range of climate impacts studies for agriculture and other sectors in the United States \citep{abatzoglou2016impact, pereira2015crop, crane2018machine, venkatappa2021impacts, zhu2019dissecting}. Daily precipitation values were extracted for each gridMET cell overlapping counties in our sample. Daily values of maximum and minimum temperature, surface radiation, wind speed, and humidity were also obtained from gridMET in order to calculate daily reference crop evapotranspiration for each grid cell. From these data, we computed the seasonal water deficit (\(D_{i,t}\)) for each grid cell by subtracting total precipitation from total reference evapotranspiration for the period May through September, which represents the typical growing season for corn and soybean in the HPA region. Seasonal water balance estimates were subsequently aggregated to the county-level to match our agricultural production data using an area-weighted average based on the proportion of the county covered by each overlapping gridMET cell. Any county-year combinations that had an extreme water balance value (lower than -200 or greater than 1200) were dropped. Figure \ref{fig:deficit-yield-hist} presents annual water deficit by year for the corn and soybean data, illustrating major drought events in the region in years such as 2000, 2002, and 2012. The importance of irrigation is clearly evident in years with large water deficits, with rainfed yields dipping substantially whereas irrigated yields are, on average, largely unaffected (Figure \ref{fig:deficit-yield-hist}).

@@

where county and year are indicated by subscripts \(i\) and \(t\), respectively, \(Y\) is crop yield, and \(D\) is seasonal water deficit. \(C_j\) is a dummy variable that takes 1 if observation \(i\) is under category \(C_j\), 0 otherwise. The observations are grouped into four categories: rainfed production, and county-years with aquifer thickness falling in the \([0\%-33.3\%)\), \([33.3\%-66.7\%)\), and \([66.7\% - 100\%]\) quantile ranges of aquifer thickness. The three irrigated production groups have about the same number of observations, which ensures statistical stability in estimating yield response to drought by category. Spline functions are used to capture the potential non-linearity of the impact of water deficit on yield in a flexible manner, with \(\sigma_j\) representing the category-specific intercept, \(\phi_{j,k}\) \(k\)th spline basis function, and \(\beta_k\) is its coefficient (see Appendix \ref{spline-basis} for further explanations on how spline-based regression works). $K$ is the total number of spline basis functions. $K = 3$ was chosen for this regression to prevent our estimated response from becoming unrealistically oscillatory while maintaining a good fit based on Bayesian Information Criterion (BIC). To control for time-invariant heterogeneity across counties and yearly shocks, county fixed effects (\(\alpha_i\)) and year fixed effects (\(\lambda_t\)) are also included and standard error estimation is clustered by county.

@@

where \(s_{i,t}\) is the share of irrigated production. The impact of average water balance is captured by \(\sum_{k}^K \eta_{k}\phi_{k}(D_{i})\), where \(\phi_{k}(\cdot)\) is the \(k\)th spline basis function and \(\eta_{k}\) is its coefficient. The impact of aquifer thickness is captured in a similar way with \(\tau_{l}(\cdot)\) and \(\beta_{l}\) being the \(l\)th spline basis function and its coefficient, respectively. $K$ and $L$ are the numbers of basis functions for $D$ and $A$, respectively. $K = 4$ and $L = 4$ were chosen to balance preventing the estimated responses from being unrealistically oscillatory and ensuring a good fit as indicated by BIC. For this regression, counties with aquifer thickness levels greater than 180 m were excluded due to the sparsity of observations in that range because our flexible functional form approach cannot reliably estimate its impact. County-level soil characteristics including sand percentage, silt percentage, and water holding capacity, are represented by \(X_i\). State-year fixed effects and error term are represented by \(\theta_{s,t}\) and \(v_{i,t}\), respectively. Unlike for our regression models of per-area yields, we do not include county fixed effects to avoid eliminating the majority of variation in the aquifer thickness variable. The vast majority of variation in aquifer thickness is cross-sectional (across counties), rather than over time. Including county fixed effects would effectively eliminate the cross-sectional variation in aquifer thickness. A formal discussion of the trade-offs associated with the choice of fixed effects is provided in Appendix \ref{county-fe}.

@@

All the raw data-sets used in this study are publicly accessible. The boundary shape file of the High Plains aquifer was obtained from the U.S. Geological Survey at \url{https://water.usgs.gov/GIS/metadata/usgswrd/XML/ds543.xml#stdorder}. County and State boundaries were obtained using the R tigris package \citep{Rtigris}. These datasets were used for Figure \ref{fig:sat-map}. Corn and soybean yield data were obtained from the U.S. Department of Agriculture's National Agricultural Statistics Service (USDA-NASS), available at \url{https://www.nass.usda.gov/Quick_Stats/}. The R tidyUSDA package \citep{RtidyUSDA} was used to automatically download the relevant data. This data is used for Figure \ref{fig:deficit-yield-hist} and the regression analysis. Weather data was obtained from the gridMET dataset \citep{Abatzoglou2013}, which is available at \url{https://www.climatologylab.org/gridmet.html}. Elevation data was obtained from the gridMET website at \url{https://www.climatologylab.org/gridmet.html}. These datasets were used to calculate water balance, which was used for Figure \ref{fig:deficit-yield-hist} and the regression analysis. The aquifer thickness data were obtained from Hydroshare \citep{haacker2023} available at \url{https://www.hydroshare.org/resource/7d925c7944244032af98c9ed20c22db6/}. This data is used for Figure \ref{fig:sat-map} and the regression analysis. Soil characteristics data was obtained from the Soil Survey Geographic Dataset, which was used for regression analysis. The R soilDB package \citep{Rsoildb} was used to download the data. Regression analyses are implemented using the R fixest package \citep{Rfixest}. Other key R packages include ggplot2 \citep{Rggplot2} for creating plots, data.table \citep{Rdatatable} for data wrangling, and sf \citep{Rsf} for spatial data operations.

\subsection{Code Availability}
All data processing and analyses were carried out using R \citep{R}. The codes used for this study are available at the following GitHub repository: \url{https://github.com/tmieno2/Drought-Production-Risk-Aquifer}.

### 2023-11-01 Writing/manuscript.tex
updates on manuscript writing (mostly on data)

\begin{document}
\maketitle
\begin{abstract}
Aquifer depletion poses a major threat to the ability of farmers, food supply chains, and rural economies globally to use groundwater as a means of adapting to climate variability and change. Empirical research has demonstrated the large differences in drought risk exposure that exist between rainfed and irrigated croplands, but previous work commonly assumes water supply for the latter is unconstrained. In this paper, we evaluate how aquifer depletion affects the resilience of irrigated crop production to drought risk using over 30 years of data on historical corn and soybean yields, production areas, and aquifer conditions for the High Plains region in the United States. We show that aquifer depletion reduces the ability of farmers to sustain irrigated crop yields and production areas in years and locations with large growing season water deficits. Our findings demonstrate that drought-related production losses on irrigated croplands increase non-linearly with aquifer depletion, highlighting the need for proactive aquifer conservation interventions to support adaptation and resilience to future increases in rainfall variability under climate change.
\end{abstract}

@@

Figure \ref{fig:yield-dif} shows the difference in yield of the second and third (highest) aquifer quantiles relative to the first (lowest) quantile with the 95\% confidence interval for corn and soybean. For corn, in years with water deficit greater than about 700 mm, the highest aquifer thickness group has statistically significantly higher yields than the lowest aquifer thickness group. At an extreme water deficit level of 950 mm (average water deficit in 2012, the most extreme drought in our historical record), the difference in yields between the smallest and largest aquifer thickness groups is 1.15 tonne/ha. Unlike corn, irrigated soybean yields are not statistically distinguishable among the three aquifer thickness groups. This is likely because soybean is not as water demanding a crop as corn, meaning that even counties with lower aquifer thickness can meet the majority of crop water demands as well yield is not a significantly binding constraint on irrigation management decisions.

% At low water deficits (below 200mm), Figure \ref{fig:yield-dif} illustrates that counties with the highest levels of aquifer thickness (i.e., 3rd quantile in Figure \ref{fig:yield-response}) experience statistically lower yields for corn and soybean than counties with the lowest level of aquifer thickness (1st quantile). In case of soybean, irrigated yields for the lowest aquifer thickness category converge to levels comparable to those obtained from rainfed production. These counter-intuitive results reflect a combination of two factors. First, in years with small water deficits there is less need for irrigation and hence irrigated and rainfed yield differences are reduced. Second, years with small or even negative water deficits are associated with wetter conditions during the growing season, which can negatively impact yield if waterlogging occurs leading to greater risks of physical crop damage (e.g., lodging of grains), delays to planting/harvesting, nutrient leaching, and restricted root growth in waterlogged soils \citep{li2019excessive}. Declines in irrigated yields could be indicative of over-irrigation by farmers, which has been shown to occur in the HPA region in wet years \citep{foster2019assessing, gibson2017case, gibson2019benchmarking}, exacerbating these risks. However, the small declines in irrigated yields at low or negative levels of water deficit may also reflect the fact that wetter years are typically cooler, reducing positive heat effects on crop growth and yield formation.

@@

\caption{The impact of water deficit and aquifer thickness on rainfed and irrigated per-area yields of corn and soybean in US High Plains. Irrigated yield responses to water deficits are dissagregated by aquifer thickness level, considering three quantile groups that capture the range of aquifer thickness values in our sample: (i) 9 m to 30.1 m; (ii) 30.1 m to 67.2 m; and (iii) 67.2 m to 215.8 m. The histograms on the top of each panel show the distribution of county-level annual water deficit values included in regressions for each crop. Total number of observations are $8,773$ and $5,977$ for corn and soybean regressions, respectively.}\label{fig:yield-response}
\end{figure}

@@

Figure \ref{fig:ir-share} shows the estimated relationship between aquifer thickness and the share of production area that is irrigated for corn and soybean after accounting for spatio-temporal variations in weather conditions and drought risk exposure across the aquifer. Figure \ref{fig:ir-share} demonstrates that aquifer thickness has meaningfully large impacts on the share of irrigated production for both corn and soybean. As aquifer thickness declines, the share of irrigated acres for corn decreases from around 0.90 (90\%) for an average county-level aquifer thickness of 100 m to 0.72 (72\%) for an average county-level aquifer thickness of 10 m (See Appendix \ref{test-dif-share} for formal statistical testings). Reductions in the share of production area under irrigation is also evident for soybean.

@@

\caption{The estimated impact of aquifer thickness on the share of agricultural production under irrigation for corn and soybean producing counties in the US High Plains (Note: shaded area represents 95\% confidence interval). The histograms on the top of each panel show the distribution of county-level aquifer thickness values included for each crop. Total number of observations are $2,523$ and $1,558$ for corn and soybean regressions, respectively.}\label{fig:ir-share}
\end{figure}

@@

Figure \ref{fig:tot-impact} and Figure \ref{fig:dif-tot-impact} also demonstrate that impact of aquifer thickness on average per-area production is highly non-linear at high seasonal water deficits. At a water deficit of 950 mm, for example, reducing from 100 m aquifer thickness to 70 m (decline of 30 m) would result in only 0.09 tonne/ha of production loss for corn. This negligible difference in average yield from the decline is due to the insensitivity of the share of irrigated acres to a decline in aquifer thickness when the initial aquifer thickness level is high as can be seen in Figure \ref{fig:ir-share}. Reducing from 70 m aquifer thickness to 40 m (decline of 30 m) would result in 0.95 tonne/ha of production loss for corn, with a further drop from 40 m aquifer thickness to 10 m (decline of 30 m) leading to 1.57 tonne/ha of production loss for corn. Similar but less pronounced effects are also observed for soybean. This non-linear impact of aquifer thickness is due to the influence of aquifer thickness on the share of irrigated corn acres (Figure \ref{fig:ir-share}), which implicitly reflects how aquifer thickness alters well yields and, in turn, farmers' ability to effectively and reliably buffer crops against water deficits during the growing season.

@@

As a proxy for the extent of aquifer depletion we use annual estimates of aquifer thickness generated by existing research \citep{haacker2016water,haacker2023}. These data provide annual estimates of the aquifer thickness at 250m resolution since 1935. Annual aquifer thickness maps were derived based on interpolation of 14,000 to 20,000 individual observations of water levels per year, along with detailed maps of both the land surface elevation and bedrock elevation of the HPA. Aquifer thickness estimates for each year were aggregated to the county level using an area-weighted average to match the resolution of agricultural production data. Aquifer thickness values were only estimated for counties with irrigated yield observations. Furthermore, during the aggregation process, any aquifer cells within a county that had a aquifer thickness value lower than 9 m were removed before aggregation to reflect the fact that aquifer thicknesses lower than this threshold are not considered to be viable for extraction and therefore it is unlikely these areas contribute groundwater for irrigation production \citep{fenichel2016measuring, haacker2016water, deines2020transitions}. Figure \ref{fig:sat-map} presents the distribution of aquifer thickness across our study area for counties included in corn and soybean analyses. Aquifer thickness levels vary substantially over space and time, which we will exploit in our analysis to quantify how aquifer thickness influences farmers' exposure and responses to drought.

These data were combined to form four sets of data for regression analysis: (1) per-acre corn yield ($8,773$ observations), (2) per-acre soybean yield ($5,977$ observations), (3) share of irrigated corn ($2,523$ observations), and (4) share of irrigated soybean ($1,558$ observations).

@@

\hypertarget{data-and-code-availability}{%
\subsection{Data Availability}\label{data-and-code-availability}}

All the codes used for this study are available at the following Github repository (\href{https://github.com/tmieno2/Drought-Production-Risk-Aquifer}{link}). All analyses were performed in R \citep{R} and all the raw data-sets used in this study are publicly accessible. The boundary shape file of the High Plains aquifer was obtained from the U.S. Geological Survey at \url{https://water.usgs.gov/GIS/metadata/usgswrd/XML/ds543.xml#stdorder}. This data is used for Figure \ref{fig:sat-map}. Corn and soybean yield data were obtained from the U.S. Department of Agriculture's National Agricultural Statistics Service (USDA-NASS), available at \url{https://www.nass.usda.gov/Quick_Stats/}. The R tidyUSDA package \citep{RtidyUSDA} was used to automatically download the relevant data. This data is used for Figure \ref{fig:deficit-yield-hist} and the regression analysis. Weather data (precipitation, temperature, and evapotranspiration) was obtained from the gridMET dataset \citep{Abatzoglou2013}, which is available at \url{https://www.climatologylab.org/gridmet.html}. This data is used for Figure \ref{fig:deficit-yield-hist} and the regression analysis. The aquifer thickness data were obtained from Hydroshare \citep{haacker2023} available at \url{https://www.hydroshare.org/resource/7d925c7944244032af98c9ed20c22db6/}. This data is used for Figure \ref{fig:sat-map} and the regression analysis. Soil characteristics data was obtained from the Soil Survey Geographic Dataset, which is used for regression analysis. The R soilDB package \citep{Rsoildb} was used to download the data. Regression analyses are implemented using the R fixest package \citep{Rfixest}. Other key R packages include ggplot2 \citep{Rggplot2} for creating plots, data.table \citep{Rdatatable} for data wrangling, and sf \citep{Rsf} for spatial data operations. The data used for the regression analysis can be generated using our codes by following the instructions available at the linked Github repository.

\subsection{Code Availability}

>>>>

\usepackage{Sweave}
\begin{document}
\maketitle
\begin{abstract}
Aquifer depletion poses a major threat to the ability of farmers, food supply chains, and rural economies globally to use groundwater as a means of adapting to climate variability and change. Empirical research has demonstrated the large differences in drought risk exposure that exist between rainfed and irrigated croplands, but previous work commonly assumes water supply for the latter is unconstrained. In this paper, we evaluate how aquifer depletion affects the resilience of irrigated crop production to drought risk using over 30 years of data on historical corn and soybean yields, production areas, and aquifer conditions for the High Plains region in the United States. We show that aquifer depletion reduces the ability of farmers to sustain irrigated crop yields and production areas in years and locations with large growing season water deficits. Our findings demonstrate that drought-related production losses on irrigated croplands increase non-linearly with aquifer depletion, highlighting the need for proactive aquifer conservation interventions to support adaptation and resilience to future increases in rainfall variability under climate change.
\end{abstract}

@@

Figure \ref{fig:yield-dif} shows the difference in yield of the second and third (highest) aquifer quantiles relative to the first (lowest) quantile with the 95\% confidence interval for corn and soybean. For corn, in years with water deficit greater than about 700 mm, the highest aquifer thickness group has statistically significantly higher yields than the lowest aquifer thickness group. At an extreme water deficit level of 950 mm (average water deficit in 2012, the most extreme drought in our historical record), the difference in yields between the smallest and largest aquifer thickness groups is 1.31 tonne/ha. Unlike corn, irrigated soybean yields are not statistically distinguishable among the three aquifer thickness groups. This is likely because soybean is not as water demanding a crop as corn, meaning that even counties with lower aquifer thickness can meet the majority of crop water demands as well yield is not a significantly binding constraint on irrigation management decisions.

@@

\caption{The impact of water deficit and aquifer thickness on rainfed and irrigated per-area yields of corn and soybean in US High Plains. Irrigated yield responses to water deficits are dissagregated by aquifer thickness level, considering three quantile groups that capture the range of aquifer thickness values in our sample: (i) 9 m to 30.1 m; (ii) 30.1 m to 67.2 m; and (iii) 67.2 m to 215.8 m. The histograms on the top of each panel show the distribution of county-level annual water deficit values included in regressions for each crop. Total number of observations are $8,814$ and $5,997$ for corn and soybean regressions, respectively.}\label{fig:yield-response}
\end{figure}

@@

Figure \ref{fig:ir-share} shows the estimated relationship between aquifer thickness and the share of production area that is irrigated for corn and soybean after accounting for spatio-temporal variations in weather conditions and drought risk exposure across the aquifer. Figure \ref{fig:ir-share} demonstrates that aquifer thickness has meaningfully large impacts on the share of irrigated production for both corn and soybean. As aquifer thickness declines, the share of irrigated acres for corn decreases from around 0.89 (89\%) for an average county-level aquifer thickness of 100 m to 0.7 (70\%) for an average county-level aquifer thickness of 10 m (See Appendix \ref{test-dif-share} for formal statistical testings). Reductions in the share of production area under irrigation is also evident for soybean.

@@

\caption{The estimated impact of aquifer thickness on the share of agricultural production under irrigation for corn and soybean producing counties in the US High Plains (Note: shaded area represents 95\% confidence interval). The histograms on the top of each panel show the distribution of county-level aquifer thickness values included for each crop. Total number of observations are $2,464$ and $1,553$ for corn and soybean regressions, respectively.}\label{fig:ir-share}
\end{figure}

@@

Figure \ref{fig:tot-impact} and Figure \ref{fig:dif-tot-impact} also demonstrate that impact of aquifer thickness on average per-area production is highly non-linear at high seasonal water deficits. At a water deficit of 950 mm, for example, reducing from 100 m aquifer thickness to 70 m (decline of 30 m) would result in only 0.15 tonne/ha of production loss for corn. This negligible difference in average yield from the decline is due to the insensitivity of the share of irrigated acres to a decline in aquifer thickness when the initial aquifer thickness level is high as can be seen in Figure \ref{fig:ir-share}. Reducing from 70 m aquifer thickness to 40 m (decline of 30 m) would result in 0.95 tonne/ha of production loss for corn, with a further drop from 40 m aquifer thickness to 10 m (decline of 30 m) leading to 1.45 tonne/ha of production loss for corn. Similar but less pronounced effects are also observed for soybean. This non-linear impact of aquifer thickness is due to the influence of aquifer thickness on the share of irrigated corn acres (Figure \ref{fig:ir-share}), which implicitly reflects how aquifer thickness alters well yields and, in turn, farmers' ability to effectively and reliably buffer crops against water deficits during the growing season.

@@

As a proxy for the extent of aquifer depletion we use annual estimates of aquifer thickness generated by existing research \citep{haacker2016water,haacker2023}. These data provide annual estimates of the aquifer thickness at 250m resolution since 1935. Annual aquifer thickness maps were derived based on interpolation of 14,000 to 20,000 individual observations of water levels per year, along with detailed maps of both the land surface elevation and bedrock elevation of the HPA. Aquifer thickness estimates for each year were aggregated to the county level using an area-weighted average to match the resolution of agricultural production data. Aquifer thickness values were only estimated for counties with irrigated yield observations. Furthermore, during the aggregation process, any county that had an aquifer thickness value lower than 9 m was removed before aggregation to reflect the fact that aquifer thicknesses lower than this threshold are not considered to be viable for extraction and therefore it is unlikely these areas contribute groundwater for irrigation production \citep{fenichel2016measuring, haacker2016water, deines2020transitions}. Figure \ref{fig:sat-map} presents the distribution of aquifer thickness across our study area for counties included in corn and soybean analyses. Aquifer thickness levels vary substantially over space and time, which we will exploit in our analysis to quantify how aquifer thickness influences farmers' exposure and responses to drought.

These data were combined to form four sets of data for regression analysis: (1) per-acre corn yield ($8,814$ observations), (2) per-acre soybean yield ($5,997$ observations), (3) share of irrigated corn ($2,464$ observations), and (4) share of irrigated soybean ($1,553$ observations).

@@

\hypertarget{data-availability}{%
\subsection{Data Availability}\label{data-availability}}

All the raw data-sets used in this study are publicly accessible. The boundary shape file of the High Plains aquifer was obtained from the U.S. Geological Survey at \url{https://water.usgs.gov/GIS/metadata/usgswrd/XML/ds543.xml#stdorder}. County and State boundaries were obtained using the R tigris package \citep{Rtigris}. These datasets were used for Figure \ref{fig:sat-map}. Corn and soybean yield data were obtained from the U.S. Department of Agriculture's National Agricultural Statistics Service (USDA-NASS), available at \url{https://www.nass.usda.gov/Quick_Stats/}. The R tidyUSDA package \citep{RtidyUSDA} was used to automatically download the relevant data. This data is used for Figure \ref{fig:deficit-yield-hist} and the regression analysis. Weather data was obtained from the gridMET dataset \citep{Abatzoglou2013}, which is available at \url{https://www.climatologylab.org/gridmet.html}. Elevation data was obtained from the gridMET website at \url{https://www.climatologylab.org/gridmet.html}. These datasets were used to calculate water balance, which was used for Figure \ref{fig:deficit-yield-hist} and the regression analysis. The aquifer thickness data were obtained from Hydroshare \citep{haacker2023} available at \url{https://www.hydroshare.org/resource/7d925c7944244032af98c9ed20c22db6/}. This data is used for Figure \ref{fig:sat-map} and the regression analysis. Soil characteristics data was obtained from the Soil Survey Geographic Dataset, which was used for regression analysis. The R soilDB package \citep{Rsoildb} was used to download the data. Regression analyses are implemented using the R fixest package \citep{Rfixest}. Other key R packages include ggplot2 \citep{Rggplot2} for creating plots, data.table \citep{Rdatatable} for data wrangling, and sf \citep{Rsf} for spatial data operations.

\subsection{Code Availability}
All data processing and analyses were carried out using R \citep{R}. The codes used for this study are available at the following GitHub repository: \url{https://github.com/tmieno2/Drought-Production-Risk-Aquifer}.
